---
title: "Experimental Design and Legacy Split-Plot Models"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Experimental Design and Legacy Split-Plot Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi = 150
)
set.seed(260818)
```


# Purpose

This vignette develops the design logic behind split-plot, split-split-plot, strip-plot, and repeated-measure experiments. It is deliberately placed before advanced mixed-model topics because the hierarchy of experimental units determines which model structures are scientifically defensible.

# 1. Learning objectives

You will learn to:

- distinguish observation, experimental unit, and sampling unit;
- identify whole-plot, subplot, and sub-subplot randomization levels;
- detect common forms of pseudoreplication;
- inspect design support and empty cells;
- understand expected mean-square logic in a classical split-plot;
- connect error strata to random effects in an LMM;
- understand when the classical balanced ANOVA becomes inadequate;
- recognize split-split and strip-plot structures.

# 2. Experimental units come before formulas

An **experimental unit** is the smallest unit that can independently receive a treatment under the randomization actually performed.

In a split-plot field trial, two treatment factors can have different experimental units.

| Level | Example | Randomized to | Consequence |
|---|---|---|---|
| block | field replicate | block | controls large-scale field variation |
| whole plot | variety / irrigation | large plot | whole-plot error for this factor |
| subplot | nitrogen / spray rate | subdivision | subplot residual for this factor |
| observation | yield sample | measured subplot | not automatically an independent replicate |

A formula does not create randomization. It represents randomization that already happened.

# 3. Inspect the split-plot dataset

```{r data}
library(mixedFlowR)
d <- mixed_data("splitplot")
head(d)
```

```{r counts}
with(d, table(block, variety))
with(d, table(variety, nitrogen))
```

The number of rows is larger than the number of whole plots because several nitrogen subplots occur inside each block-by-variety whole plot.

# 4. Design audit and hierarchy

```{r audit}
audit <- mixed_design_audit(
  d,
  response = "yield",
  block    = "block",
  whole    = "variety",
  sub      = "nitrogen"
)
audit
```

```{r hierarchy}
mixed_hierarchy(
  d,
  block = "block",
  whole = "variety",
  sub   = "nitrogen"
)
```

```{r balance}
support <- mixed_balance(d, c("block", "variety", "nitrogen"))
support
```

## What the audit can and cannot know

The software can inspect the declared columns and observed support. It cannot reconstruct undocumented field operations. If the whole-plot factor was physically applied at another scale, the analyst must provide that information.

# 5. Pseudoreplication: a practical diagnostic

Create the whole-plot identifier explicitly.

```{r wholeplot-id}
d$whole_plot <- interaction(d$block, d$variety, drop = TRUE)
length(unique(d$whole_plot))
nrow(d)
```

Screen treatment assignment at that cluster level.

```{r pseudo}
mixed_pseudorep(
  d,
  response  = "yield",
  cluster   = "whole_plot",
  treatment = "variety"
)
```

A common error is to count every subplot as an independent replicate of variety. The true replication for variety is at the whole-plot level.

# 6. Classical split-plot decomposition

For a textbook classical ANOVA, represent nitrogen as a categorical subplot factor.

```{r factor-dose}
d$nitrogen_f <- factor(d$nitrogen)
```

Fit the legacy analysis.

```{r legacy}
legacy <- fit_legacy_splitplot(
  d,
  response = "yield",
  block    = "block",
  whole    = "variety",
  sub      = "nitrogen_f"
)
legacy
```

Extract the flattened ANOVA table.

```{r legacy-anova}
legacy_tab <- splitplot_anova(legacy)
legacy_tab
```

Display the expected mean-square teaching table.

```{r ems}
ems <- splitplot_ems(
  d,
  block = "block",
  whole = "variety",
  sub   = "nitrogen_f"
)
ems
```

# 7. How to read the error strata

For a balanced split-plot, the conceptual rule is:

- whole-plot factor: tested against variation among whole plots within blocks;
- subplot factor: tested against subplot residual variation;
- whole-by-sub interaction: also uses subplot-level variation in the classical decomposition.

This is why ordinary two-way ANOVA is not a valid substitute for split-plot ANOVA.

# 8. Transition to a mixed model

```{r mixed-equivalent}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit_factor <- mixed_fit(
    yield ~ variety * nitrogen_f +
      (1 | block) +
      (1 | block:variety),
    data = d
  )

  fit_factor
}
```

The model can be read as:

- fixed population effects for variety, nitrogen, and their interaction;
- random block-to-block variation;
- random whole-plot variation through block-by-variety;
- subplot residual variation.

Compare directly.

```{r compare}
if (requireNamespace("lme4", quietly = TRUE)) {
  compare_legacy_mixed(
    d,
    "yield",
    "block",
    "variety",
    "nitrogen_f"
  )
}
```

**Interpretation.** Close agreement is expected only when the parameterizations and balanced-design assumptions are comparable. The point is to validate the randomization logic, not to force exact equality of every summary statistic under different inferential conventions.

# 9. Quantitative subplot treatment: a different question

A classical factor ANOVA asks whether dose levels differ. A regression-oriented mixed model asks how yield changes along a quantitative nitrogen scale.

```{r numeric-dose}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit_numeric <- mixed_fit(
    yield ~ variety * poly(nitrogen, 2) +
      (1 | block) +
      (1 | block:variety),
    data = d
  )
}
```

```{r numeric-curve}
if (
  exists("fit_numeric") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  mixed_plot(
    fit_numeric,
    type      = "curve",
    x         = "nitrogen",
    by        = "variety",
    values    = seq(min(d$nitrogen), max(d$nitrogen), length.out = 61),
    show_data = TRUE
  )
}
```

This approach is often more useful when dose has a mechanistic or agronomic continuum.

# 10. Deliberately create an incomplete design

Remove one subplot observation.

```{r incomplete}
d_incomplete <- d[-1, ]
mixed_balance(d_incomplete, c("block", "variety", "nitrogen"))
```

The classical balanced decomposition is now less attractive. A mixed model can accommodate missing observations under its likelihood assumptions, but the missingness mechanism and reduced information must still be considered.

# 11. Split-split-plot design

A split-split plot introduces a third randomization scale. To illustrate the API, construct a teaching factor `stage`. This is only a software demonstration; a real split-split factor must come from genuine randomization.

```{r splitsplit-data}
d_ss <- d
d_ss$stage <- factor(rep(c("S1", "S2"), length.out = nrow(d_ss)))
```

```{r splitsplit-fit}
ss <- fit_legacy_splitsplit(
  d_ss,
  response = "yield",
  block    = "block",
  whole    = "variety",
  sub      = "nitrogen_f",
  subsub   = "stage"
)
ss
```

**Important.** Never manufacture a sub-subplot factor in a real analysis just to fit a more complex model. The code above is strictly a syntax demonstration.

# 12. Strip-plot design

Strip plots arise when two factors are randomized in perpendicular or intersecting strips, creating distinct error terms for each strip factor and their intersection.

For API demonstration:

```{r strip-data}
d_strip <- d
d_strip$A <- factor(d_strip$variety)
d_strip$B <- factor(d_strip$nitrogen)
```

```{r strip-fit}
strip <- fit_legacy_stripplot(
  d_strip,
  response = "yield",
  block    = "block",
  strip_a  = "A",
  strip_b  = "B"
)
strip
```

A true strip-plot analysis requires the field layout to satisfy this randomization structure.

# 13. Repeated measures are not automatically split plots

Repeated measurements share the same observational unit over time. They create within-unit dependence. Sometimes repeated measures occur inside a split-plot experiment, but the two structures should not be conflated.

```{r repeated-data}
z <- mixed_data("longitudinal")
head(z)
```

The repeated subject can be declared in a design audit.

```{r repeated-audit}
mixed_design_audit(
  z,
  response = "height",
  subject  = "subject",
  time     = "time"
)
```

Later, `mixed_ar1()`, `mixed_car1()`, and `mixed_heterogeneity()` will model residual dependence that remains after the fixed and random effects.

# 14. Three common design failures

## Failure 1: wrong replication count

**Mistake:** treating subplots as independent variety replicates.

**Consequence:** standard errors for the whole-plot factor can be severely underestimated.

**Prevention:** explicitly define the whole plot and random structure.

## Failure 2: random effect chosen from software convention

**Mistake:** adding `(1|block)` because it is common, while omitting the whole-plot experimental unit.

**Consequence:** the model may not represent the randomization.

**Prevention:** derive grouping factors from field operations.

## Failure 3: changing a quantitative factor to categorical only to obtain letters

**Mistake:** reducing a dose-response question to many pairwise tests.

**Consequence:** loss of shape information and potentially lower power.

**Prevention:** use `mixed_trend()`, `mixed_curve()`, and `mixed_optimum()` when the quantitative scale is meaningful.

# 15. Reporting a split-plot analysis

A methods section should report at least:

1. block structure and number of blocks;
2. whole-plot experimental unit;
3. subplot experimental unit;
4. treatment allocation and randomization;
5. whether subplot treatment was modeled categorically or quantitatively;
6. random-effects structure;
7. estimation method;
8. denominator-DF approach for fixed-effect tests when used;
9. diagnostics and any sensitivity analysis;
10. estimates with uncertainty.

# 16. Practice exercise: from wrong to correct

Fit the intentionally incomplete model below and compare its random-effect structure with the design-correct model.

```{r wrong-model}
if (requireNamespace("lme4", quietly = TRUE)) {
  wrong <- mixed_fit(
    yield ~ variety * nitrogen_f + (1 | block),
    data = d
  )

  correct <- mixed_fit(
    yield ~ variety * nitrogen_f +
      (1 | block) +
      (1 | block:variety),
    data = d
  )
}
```

Do not decide between them only from AIC. Explain which randomization level is missing from `wrong`.

# 17. Key take-home messages

- Experimental-unit hierarchy is a property of the experiment, not the software.
- Whole-plot and subplot treatments have different error scales.
- Legacy ANOVA is highly valuable for teaching and balanced-design auditing.
- Mixed models generalize the representation but do not remove the need to understand randomization.
- Quantitative subplot treatments deserve regression-oriented interpretation when scientifically appropriate.

# 18. Next vignette

Continue with `03-random-effects-and-shrinkage.Rmd` to understand the statistical meaning of random intercepts, random slopes, nesting, crossing, BLUPs, and partial pooling.
