---
title: "GLMM Distributions: Counts, Incidence, Proportions, and Zero Inflation"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{GLMM Distributions: Counts, Incidence, Proportions, and Zero Inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi = 150
)
set.seed(260818)
```


# Purpose

This vignette extends the mixed-model framework to non-Gaussian responses. The focus is not on memorizing distributions. The objective is to connect the measurement process to an appropriate conditional distribution, link function, random structure, and diagnostic workflow.

`mixedFlowR` uses `lme4` and `glmmTMB` as principal GLMM engines, with `glmmTMB` providing especially broad support for negative-binomial, beta, zero-inflated, and dispersion models [@brooks2017].

# 1. Learning objectives

You will learn to:

- distinguish binary/binomial, count, and continuous-proportion outcomes;
- understand the role of the link function;
- fit binomial, Poisson, negative-binomial, beta, and zero-inflated mixed models;
- recognize overdispersion and excess-zero questions;
- use simulation-based residual diagnostics when `DHARMa` is available;
- compare candidate models without automatic model selection;
- report effects on both link and response scales when useful.

# 2. A practical family map

| Response | Typical family | Link | Common complication |
|---|---|---|---|
| successes out of trials | binomial | logit | overdispersion / grouping |
| insect count | Poisson | log | overdispersion |
| overdispersed count | negative binomial | log | zero inflation |
| proportion strictly in `(0,1)` | beta | logit | boundary values |
| structural + sampling zeros | zero-inflated family | family-specific | distinguish zero mechanisms |

The distribution should be motivated by the response scale and data-generating process, not by which model gives the smallest AIC.

# 3. Binomial incidence example

Load the disease-incidence teaching data.

```{r binomial-data}
library(mixedFlowR)
b <- mixed_data("binomial")
head(b)
```

Observed treatment totals:

```{r binomial-summary}
aggregate(cbind(diseased, healthy, total) ~ treatment, data = b, FUN = sum)
```

Observed incidence:

```{r incidence}
inc_obs <- aggregate(cbind(diseased, total) ~ treatment, data = b, FUN = sum)
inc_obs$incidence <- inc_obs$diseased / inc_obs$total
inc_obs
```

Fit a grouped binomial GLMM.

```{r binomial-fit}
if (requireNamespace("lme4", quietly = TRUE)) {
  bin_fit <- mixed_binomial(
    data      = b,
    successes = "diseased",
    failures  = "healthy",
    fixed     = "treatment",
    group     = "block",
    engine    = "lme4"
  )

  bin_fit
}
```

Obtain model-based means on the response scale.

```{r binomial-emmeans}
if (
  exists("bin_fit") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  mixed_emmeans(
    bin_fit,
    specs = ~ treatment,
    type = "response"
  )
}
```

**Interpretation.** On the response scale, estimates are probabilities rather than logits. Report both the modeled treatment contrast and a response-scale effect that readers can interpret biologically.

# 4. Count data: Poisson as a starting model

```{r count-data}
cnt <- mixed_data("counts")
head(cnt)
```

Observed mean and zero fraction:

```{r count-summary}
aggregate(count ~ treatment, data = cnt, FUN = mean)
aggregate(I(count == 0) ~ treatment, data = cnt, FUN = mean)
```

Fit a Poisson GLMM.

```{r poisson-fit}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  pois <- mixed_count(
    count ~ treatment * time + (1 | plot),
    data   = cnt,
    family = "poisson",
    engine = "glmmTMB"
  )

  pois
}
```

The Poisson assumption equates conditional mean and conditional variance. Biological count data often violate this restriction.

# 5. Negative-binomial alternatives

Fit NB2.

```{r nb2-fit}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  nb2 <- mixed_count(
    count ~ treatment * time + (1 | plot),
    data   = cnt,
    family = "nbinom2"
  )
}
```

Fit NB1.

```{r nb1-fit}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  nb1 <- mixed_count(
    count ~ treatment * time + (1 | plot),
    data   = cnt,
    family = "nbinom1"
  )
}
```

The parameterizations differ in how variance increases with the conditional mean. Use diagnostics and subject-matter knowledge rather than assuming one is universally preferable.

# 6. Simulation-based diagnostics with DHARMa

```{r dharma-nb2}
if (
  exists("nb2") &&
  requireNamespace("DHARMa", quietly = TRUE)
) {
  nb_diag <- mixed_diagnose(
    nb2,
    simulate = TRUE,
    nsim = 500
  )

  nb_diag
}
```

Simulation-based residuals are particularly valuable for discrete outcomes because raw residual patterns are often difficult to interpret directly.

Look for:

- systematic residual trends;
- over- or underdispersion;
- excess zeros;
- remaining temporal or spatial autocorrelation;
- influential clusters.

# 7. Zero inflation is a process hypothesis

Fit a zero-inflated negative-binomial model.

```{r zi-fit}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  zi <- mixed_zero_inflated(
    count ~ treatment * time + (1 | plot),
    data        = cnt,
    family      = glmmTMB::nbinom2,
    ziformula   = ~ treatment,
    dispformula = ~ 1
  )

  zi
}
```

**Important interpretation.** A zero-inflated model assumes an additional mechanism capable of producing zeros beyond the conditional count distribution. The mere presence of many zeros does not prove that such a process exists.

# 8. Compare count candidates

Information criteria can summarize relative penalized likelihood.

```{r count-compare}
if (exists("pois") && exists("nb1") && exists("nb2")) {
  mixed_compare(
    Poisson = pois,
    NB1     = nb1,
    NB2     = nb2,
    method  = "information"
  )
}
```

If `zi` is available:

```{r zi-compare}
if (exists("nb2") && exists("zi")) {
  mixed_compare(
    NB2 = nb2,
    ZINB = zi,
    method = "information"
  )
}
```

Do not turn this table into an automatic winner selector. Ask whether the more complex model solves a diagnosed problem and represents a plausible process.

# 9. Beta mixed model for continuous proportions

```{r beta-data}
be <- mixed_data("beta")
summary(be$severity)
range(be$severity)
```

The teaching data lie strictly between zero and one, so a beta response is admissible.

```{r beta-fit}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  beta_fit <- mixed_beta(
    severity ~ treatment + (1 | block),
    data = be,
    link = "logit"
  )

  beta_fit
}
```

Response-scale marginal means:

```{r beta-emm}
if (
  exists("beta_fit") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  mixed_emmeans(
    beta_fit,
    specs = ~ treatment,
    type = "response"
  )
}
```

# 10. Boundary values: a deliberate failure

A standard beta distribution has support in the open interval `(0,1)`.

```{r beta-boundary, eval=FALSE}
be_bad <- be
be_bad$severity[1] <- 0
mixed_beta(severity ~ treatment + (1 | block), be_bad)
```

This example is not evaluated because it is intended to demonstrate an error. Do not silently add a small constant to every observation without a documented modeling rationale.

# 11. General GLMM wrapper

The lower-level wrapper is useful when a family is directly supported by the selected engine.

```{r generic-glmm}
if (requireNamespace("lme4", quietly = TRUE)) {
  generic_bin <- mixed_glmm(
    cbind(diseased, healthy) ~ treatment + (1 | block),
    data   = b,
    family = stats::binomial(),
    engine = "lme4"
  )
}
```

# 12. Link-scale versus response-scale interpretation

With a log link, additive effects on the linear predictor become multiplicative effects on the mean response.

With a logit link, additive effects are on log-odds. A response-scale probability or risk difference may be easier to communicate.

The reporting workflow should therefore distinguish:

- coefficient scale;
- transformed response scale;
- contrast scale;
- absolute versus relative effect.

# 13. Plot observed data and fitted structure

For counts, a raw-data panel can reveal discreteness and zero structure.

```{r count-raw-plot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(cnt, ggplot2::aes(time, count, colour = treatment)) +
    ggplot2::geom_jitter(width = 0.08, height = 0, alpha = 0.55) +
    ggplot2::stat_summary(fun = mean, geom = "line", ggplot2::aes(group = treatment), linewidth = 0.8) +
    ggplot2::labs(
      x = "Time",
      y = "Pest count",
      colour = "Treatment",
      title = "Observed count trajectories"
    ) +
    ggplot2::theme_minimal(base_size = 11)
}
```

**How to read the figure.** The plotted treatment means are descriptive. The mixed model additionally accounts for repeated plot structure, conditional distribution, and any modeled zero or dispersion process.

# 14. A practical model-building sequence for counts

A defensible sequence is:

1. specify the fixed and random structure from the design;
2. fit a Poisson baseline if scientifically reasonable;
3. inspect dispersion and residual diagnostics;
4. move to a negative-binomial family if overdispersion is material;
5. evaluate zero inflation only if diagnostics and biology support a second zero mechanism;
6. add residual temporal/spatial structure if dependence remains;
7. compare conclusions across reasonable candidates.

# 15. Reporting checklist for GLMMs

Report:

- response definition and denominator when relevant;
- distribution family;
- link function;
- random effects;
- zero-inflation formula, if fitted;
- dispersion formula, if fitted;
- diagnostics;
- effect scale used in interpretation;
- estimates with confidence intervals;
- multiplicity adjustment for multiple contrasts.

# 16. Practice exercises

## Exercise 1: Poisson versus NB2

Fit both models and inspect `mixed_diagnose(..., simulate=TRUE)` if `DHARMa` is available. Does the more flexible family solve a visible residual problem?

## Exercise 2: zero-inflation formula

Compare `ziformula = ~ 1` with `ziformula = ~ treatment`. Explain the scientific claim implied by allowing the structural-zero probability to vary among treatments.

```{r exercise-zi}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  zi_intercept <- mixed_zero_inflated(
    count ~ treatment * time + (1 | plot),
    data = cnt,
    family = glmmTMB::nbinom2,
    ziformula = ~ 1
  )
}
```

## Exercise 3: beta link

Fit beta models with logit and probit links. Compare fitted conclusions on the response scale rather than coefficient magnitude alone.

```{r exercise-beta}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  beta_logit <- mixed_beta(severity ~ treatment + (1 | block), be, "logit")
  beta_probit <- mixed_beta(severity ~ treatment + (1 | block), be, "probit")
}
```

# 17. Key take-home messages

- Distribution choice starts from the response-generating process.
- Random effects account for grouping but do not automatically solve overdispersion or residual dependence.
- Negative-binomial and zero-inflated models answer different questions.
- Beta models require support strictly inside `(0,1)` unless a boundary-capable alternative is used.
- Simulation-based diagnostics are highly useful for GLMMs.
- A lower information criterion alone is not a scientific argument for complexity.

# 18. What comes next

The next stage begins with `05-residual-covariance.Rmd`, where we separate group-level random effects from residual temporal correlation and heteroscedasticity.

# References

`lme4` provides a foundational implementation of linear and generalized linear mixed models [@bates2015]. `glmmTMB` extends this framework with flexible families, zero-inflation, and related model components [@brooks2017].
