---
title: "Robust Mixed Models and Sensitivity Analysis"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Robust Mixed Models and Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

Robust analysis is most useful when it is framed as a sensitivity analysis: would the scientific conclusion change if influential observations or clusters received less leverage under a robust estimator?

The `robustlmm` backend implements robust linear mixed-model estimation [@koller2016]. `mixedFlowR` also supports influence screens, classical-versus-robust comparisons, contamination summaries, and cluster-robust covariance through optional backends.

# 1. Learning objectives

You will learn to:

- fit a robust Gaussian mixed model;
- compare classical and robust fixed-effect estimates;
- identify residual and cluster-level sensitivity;
- distinguish influence from data error;
- use cluster-robust covariance as another sensitivity tool;
- construct a transparent contamination analysis;
- report robust results without silently replacing the primary model.

# 2. Baseline longitudinal data

```{r data}
library(mixedFlowR)
z <- mixed_data("longitudinal")
```

Fit a classical LMM.

```{r classical}
if (requireNamespace("lme4", quietly = TRUE)) {
  classical <- mixed_fit(
    height ~ treatment * time + (1 | subject),
    data = z
  )
}
```

# 3. Fit a robust LMM

```{r robust}
if (requireNamespace("robustlmm", quietly = TRUE)) {
  robust <- mixed_robust(
    height ~ treatment * time + (1 | subject),
    data = z
  )

  robust
}
```

The robust estimator is not a generic replacement for every mixed model. It targets sensitivity to contamination within the supported Gaussian LMM framework.

# 4. Compare classical and robust fixed effects

```{r compare}
if (exists("classical") && exists("robust")) {
  robust_shift <- mixed_compare_robust(
    classical,
    robust
  )

  robust_shift
}
```

Important columns include the classical estimate, robust estimate, absolute shift, and relative shift where defined.

**Interpretation.** A small shift supports stability under this robustification. A large shift signals sensitivity and should trigger data review and substantive investigation.

# 5. Create a deliberately contaminated copy

Add an extreme teaching perturbation to one observation.

```{r contaminate}
z_cont <- z
z_cont$height[1] <- z_cont$height[1] + 30
```

Plot the perturbation.

```{r contamination-plot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(z_cont, ggplot2::aes(time, height, group = subject, colour = treatment)) +
    ggplot2::geom_line(alpha = 0.25) +
    ggplot2::geom_point(alpha = 0.5) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::labs(title = "Teaching contamination scenario")
}
```

# 6. Classical fit under contamination

```{r classical-cont}
if (requireNamespace("lme4", quietly = TRUE)) {
  classical_cont <- mixed_fit(
    height ~ treatment * time + (1 | subject),
    data = z_cont
  )
}
```

# 7. Robust fit under contamination

```{r robust-cont}
if (requireNamespace("robustlmm", quietly = TRUE)) {
  robust_cont <- mixed_robust(
    height ~ treatment * time + (1 | subject),
    data = z_cont
  )
}
```

```{r compare-cont}
if (exists("classical_cont") && exists("robust_cont")) {
  mixed_compare_robust(
    classical_cont,
    robust_cont
  )
}
```

This is the type of scenario in which classical-versus-robust comparison becomes instructive.

# 8. Observation-level robust influence screen

```{r influence-observation}
if (exists("classical_cont")) {
  obs_screen <- mixed_robust_influence(
    classical_cont
  )

  head(obs_screen[order(-obs_screen$abs_robust_z), ])
}
```

# 9. Cluster-level screen

```{r influence-cluster}
if (exists("classical_cont")) {
  cluster_screen <- mixed_robust_influence(
    classical_cont,
    cluster = "subject"
  )

  cluster_screen
}
```

Cluster aggregation helps identify whether sensitivity is concentrated in one experimental unit rather than one isolated measurement.

# 10. Contamination flagging

```{r contamination-screen}
if (exists("classical_cont")) {
  flags <- mixed_contamination(
    classical_cont,
    cluster   = "subject",
    threshold = 2.5
  )

  flags
}
```

Rows are flagged for **review**, never automatically deleted.

# 11. What to do after a flag

A defensible sequence is:

1. check the original field sheet, sensor file, or laboratory record;
2. verify units and data entry;
3. identify whether the value is physically plausible;
4. determine whether the observation belongs to a special but valid condition;
5. compare primary and sensitivity analyses;
6. document any exclusion criterion independently of the desired statistical result.

# 12. Cluster-robust covariance

`clubSandwich` can provide cluster-robust covariance estimates for supported models.

```{r cr2}
if (
  exists("classical") &&
  requireNamespace("clubSandwich", quietly = TRUE)
) {
  V_CR2 <- mixed_robust_vcov(
    classical,
    cluster = "subject",
    type    = "CR2"
  )

  V_CR2
}
```

CR2 changes the covariance estimator for fixed effects. It is not the same method as fitting a robust mixed model with `robustlmm`.

# 13. Group-deletion influence for lme4

```{r influence-me}
if (
  exists("classical") &&
  requireNamespace("influence.ME", quietly = TRUE)
) {
  infl <- mixed_influence(
    classical,
    group = "subject"
  )

  infl
}
```

This evaluates sensitivity to deleting whole grouping units and complements residual-based screens.

# 14. Robust analysis does not solve model misspecification

If residuals show a strong nonlinear time trend, omitted treatment interaction, or serial dependence, downweighting unusual observations does not correct the mean or covariance model.

Robustness should therefore come **after** the core design and structural modeling decisions.

# 15. A three-model sensitivity panel

A useful report can compare:

- classical LMM estimate;
- robust LMM estimate;
- classical estimate with cluster-robust covariance.

These methods address different vulnerabilities and should not be described as interchangeable.

# 16. Robust estimate table

```{r robust-table}
if (exists("robust_cont")) {
  mixed_table(
    robust_cont,
    component = "fixed",
    format = "markdown",
    caption = "Robust mixed-model fixed-effect estimates"
  )
}
```

# 17. A stricter contamination threshold

```{r strict-threshold}
if (exists("classical_cont")) {
  mixed_contamination(
    classical_cont,
    cluster = "subject",
    threshold = 3
  )
}
```

The threshold is a screening setting, not an objective truth boundary.

# 18. Reporting template

A transparent sensitivity paragraph should state:

1. why robust analysis was performed;
2. estimator/backend used;
3. whether influential observations or clusters were identified;
4. how classical and robust estimates differed;
5. whether the primary biological conclusion changed;
6. whether any observation was excluded and why;
7. whether cluster-robust covariance altered uncertainty materially.

# 19. Exercises

## Exercise 1

Move the contamination from one observation to all measurements of one subject. Compare the change in classical and robust estimates.

## Exercise 2

Compare `CR1` and `CR2` covariance estimates when `clubSandwich` is available.

```{r exercise-cr}
if (
  exists("classical") &&
  requireNamespace("clubSandwich", quietly = TRUE)
) {
  V_CR1 <- mixed_robust_vcov(classical, "subject", "CR1")
  V_CR2 <- mixed_robust_vcov(classical, "subject", "CR2")
}
```

## Exercise 3

Use `mixed_contamination()` on the uncontaminated dataset. Explain why a flag can still represent a valid extreme observation rather than an error.

# 20. Key take-home messages

- Robust estimation is a sensitivity tool, not a license to ignore model structure.
- Influence and contamination flags trigger review, not automatic deletion.
- Classical and robust estimates should be compared transparently.
- Cluster-robust covariance and robust mixed-model fitting address different inferential vulnerabilities.
- Exclusions require an independently defensible scientific or data-quality reason.

# 21. What comes next

Stage 3 begins with `07-bootstrap-inference.Rmd`, which develops interval estimation and parametric-bootstrap comparisons for mixed models.

# References

The robust LMM module builds on the robust scoring-equation framework implemented in `robustlmm` [@koller2016].
