---
title: "Bootstrap Inference for Mixed Models"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bootstrap Inference for Mixed Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

Mixed-model uncertainty is not always well summarized by a symmetric Wald interval. Variance components are constrained, random-effect structures can be close to boundaries, and small-sample fixed-effect tests can have nonstandard behavior.

This vignette introduces parametric bootstrap inference, variance-component bootstrap, prediction intervals, BCa/wild options where supported, and parametric-bootstrap model comparison.

# 1. Learning objectives

You will learn to:

- distinguish a bootstrap distribution from the original sampling model;
- obtain bootstrap intervals for fixed effects;
- bootstrap variance components;
- obtain bootstrap prediction intervals;
- understand population versus conditional predictions;
- use BCa or wild bootstrap when `confintROB` supports the fitted object;
- choose a practical number of replicates for teaching versus scientific analysis;
- report seeds, replication counts, interval type, and resampling method.

# 2. Baseline model

```{r data}
library(mixedFlowR)
z <- mixed_data("longitudinal")
```

```{r model}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit <- mixed_fit(
    height ~ treatment * time + (1 | subject),
    data = z
  )
}
```

# 3. Why bootstrap?

A model-based bootstrap asks what estimates would be obtained if new datasets were generated repeatedly under the fitted model and the analysis were repeated.

The workflow is:

1. fit the model;
2. simulate a new response under the fitted model;
3. refit the model;
4. recompute the statistic;
5. repeat many times;
6. summarize the empirical distribution.

This procedure makes the computational burden explicit: 4,999 bootstrap replicates can mean roughly 4,999 model refits.

# 4. Teaching-scale fixed-effect bootstrap

Use a deliberately small number of replications for vignette execution.

```{r bootstrap-demo}
if (exists("fit")) {
  boot_demo <- mixed_boot_fixed(
    fit,
    B = 49,
    level = 0.95,
    seed = 260818
  )

  boot_demo
}
```

The small `B` is suitable for learning the API, not for final scientific inference.

# 5. Production-scale fixed-effect bootstrap

```{r bootstrap-production, eval=FALSE}
boot_fixed <- mixed_boot_fixed(
  fit,
  B = 4999,
  level = 0.95,
  seed = 260818
)
```

Why 4,999 rather than a universal magic number? Because Monte Carlo precision depends on the tail probability being estimated, computational budget, and the stability required for the scientific conclusion. Report the selected `B` explicitly.

# 6. Bootstrap interval table

```{r boot-table}
if (exists("boot_demo")) {
  mixed_table(
    boot_demo,
    component = "bootstrap",
    format = "markdown",
    caption = "Teaching-scale parametric bootstrap intervals"
  )
}
```

**How to read the table.** Compare the original coefficient estimate with the empirical bootstrap interval. Look especially for asymmetry or unexpectedly wide uncertainty.

# 7. Visualize the bootstrap distribution

```{r boot-plot}
if (exists("boot_demo")) {
  mixed_plot(
    boot_demo,
    type = "bootstrap"
  )
}
```

The distribution shown here is the empirical distribution of the requested bootstrap statistic across refitted simulated datasets.

# 8. Variance-component bootstrap

Variance components are nonnegative and often have skewed uncertainty.

```{r variance-demo}
if (exists("fit")) {
  boot_var_demo <- mixed_boot_variance(
    fit,
    B = 49,
    seed = 260818
  )

  boot_var_demo
}
```

Production example:

```{r variance-production, eval=FALSE}
boot_var <- mixed_boot_variance(
  fit,
  B = 4999,
  seed = 260818
)
```

# 9. Random-slope variance

```{r slope-fit}
if (requireNamespace("lme4", quietly = TRUE)) {
  rs <- mixed_random_slopes(
    z,
    response = "height",
    fixed = "treatment * time",
    slope = "time",
    group = "subject"
  )
}
```

```{r slope-bootstrap}
if (exists("rs")) {
  rs_var_demo <- mixed_boot_variance(
    rs,
    B = 49,
    seed = 18
  )
}
```

A broad distribution near zero is a useful signal that the random-slope variance is weakly identified.

# 10. Bootstrap predictions

Create a prediction grid at the final time for both treatments.

```{r pred-grid}
new_height <- data.frame(
  treatment = factor(c("T1", "T2"), levels = levels(z$treatment)),
  time = c(max(z$time), max(z$time)),
  subject = z$subject[1]
)
new_height
```

Population-level predictions omit subject-specific random effects.

```{r pred-population}
if (exists("fit")) {
  pred_pop_demo <- mixed_boot_predict(
    fit,
    newdata = new_height,
    B = 49,
    re.form = NA,
    seed = 260818
  )

  pred_pop_demo
}
```

Conditional predictions include random effects for an existing subject.

```{r pred-conditional}
if (exists("fit")) {
  pred_cond_demo <- mixed_boot_predict(
    fit,
    newdata = new_height,
    B = 49,
    re.form = NULL,
    seed = 260818
  )

  pred_cond_demo
}
```

**Interpretation.** Population and conditional predictions answer different questions. The first concerns the fixed population trajectory; the second conditions on estimated subject effects.

# 11. BCa intervals

`mixed_boot()` delegates BCa support to `confintROB` for supported model classes.

```{r bca, eval=FALSE}
if (
  requireNamespace("confintROB", quietly = TRUE) &&
  exists("fit")
) {
  boot_bca <- mixed_boot(
    fit,
    statistic = "fixed",
    B = 1999,
    method = "parametric",
    interval = "BCa",
    cluster = "subject",
    seed = 260818
  )
}
```

This is not evaluated during vignette build because it can be computationally expensive and depends on an optional backend.

# 12. Wild bootstrap

A wild bootstrap can be useful when the resampling scheme is intended to be less reliant on homogeneous residual variance assumptions and the backend supports it.

```{r wild, eval=FALSE}
if (
  requireNamespace("confintROB", quietly = TRUE) &&
  exists("fit")
) {
  boot_wild <- mixed_boot(
    fit,
    statistic = "fixed",
    B = 1999,
    method = "wild",
    interval = "percentile",
    cluster = "subject",
    seed = 260818
  )
}
```

Always describe the actual implementation and cluster definition rather than reporting only the word "bootstrap."

# 13. Bootstrap for a robust fit

```{r robust-fit}
if (requireNamespace("robustlmm", quietly = TRUE)) {
  robust <- mixed_robust(
    height ~ treatment * time + (1 | subject),
    data = z
  )
}
```

```{r robust-boot, eval=FALSE}
if (
  exists("robust") &&
  requireNamespace("confintROB", quietly = TRUE)
) {
  robust_ci <- mixed_boot(
    robust,
    statistic = "fixed",
    B = 1999,
    method = "parametric",
    interval = "BCa",
    cluster = "subject",
    seed = 260818
  )
}
```

# 14. Parametric-bootstrap model comparison

`pbkrtest` can compare two nested `lmer` models through parametric bootstrap.

```{r comparison-models}
if (requireNamespace("lme4", quietly = TRUE)) {
  small <- mixed_fit(
    height ~ time + (1 | subject),
    data = z
  )
  large <- mixed_fit(
    height ~ treatment + time + (1 | subject),
    data = z
  )
}
```

Vignette-scale demonstration:

```{r pb-compare-demo}
if (
  exists("small") && exists("large") &&
  requireNamespace("pbkrtest", quietly = TRUE)
) {
  pb_cmp_demo <- mixed_compare(
    small,
    large,
    method = "parametric-bootstrap",
    nsim = 99,
    seed = 260818
  )
  pb_cmp_demo
}
```

Production version:

```{r pb-compare-production, eval=FALSE}
pb_cmp <- mixed_compare(
  small,
  large,
  method = "parametric-bootstrap",
  nsim = 4999,
  seed = 260818
)
```

# 15. Monte Carlo error matters

If the bootstrap p-value or interval endpoint is near a decision threshold, rerun with a larger simulation count. A stable conclusion should not depend strongly on changing `B` from 999 to 4,999.

A useful reproducibility check is to run the same analysis with two documented seeds and verify that substantive conclusions do not change.

# 16. What bootstrap does not fix

Bootstrap inference does not rescue:

- the wrong experimental unit;
- an inappropriate response distribution;
- unmodeled spatial/temporal dependence;
- nonconvergence;
- a scientifically implausible mean structure;
- severe extrapolation beyond the data.

Fix structural model problems before resampling them thousands of times.

# 17. Reporting checklist

Report:

- statistic bootstrapped;
- parametric or wild method;
- percentile or BCa interval;
- number of replicates;
- confidence level;
- cluster definition if used;
- seed;
- prediction conditioning (`re.form`) if prediction intervals are reported;
- any failed refits or convergence issues.

# 18. Exercises

1. Compare bootstrap intervals with Wald intervals for the same coefficient.
2. Increase `B` from 49 to 199 and observe Monte Carlo stabilization.
3. Bootstrap a random-slope variance and inspect boundary behavior.
4. Compare population and conditional prediction intervals.

# 19. Key take-home messages

- Bootstrap inference is computationally intensive because models are refitted repeatedly.
- Use small `B` only for teaching and syntax checks.
- Variance components often benefit from asymmetric interval methods.
- Population and conditional predictions are different estimands.
- Always report the resampling method, replication count, interval type, and seed.

# 20. Next vignette

Continue with `08-gamlss-mixed.Rmd` for distributional regression beyond a single mean parameter.
