---
title: "Bayesian Multilevel Models: Priors, Posterior Inference, and Prediction"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Bayesian Multilevel Models: Priors, Posterior Inference, and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

This vignette introduces the Bayesian workflow implemented through `brms`, which fits generalized, nonlinear, multivariate, and distributional multilevel models using Stan [@burkner2017].

The Bayesian module is intentionally more than a wrapper for obtaining credible intervals. A coherent workflow includes prior specification, prior predictive assessment, posterior diagnostics, posterior predictive checks, and predictive model evaluation.

Because compiling and sampling Stan models can be expensive, **Bayesian fitting chunks in this vignette are shown with `eval=FALSE` by design**. This keeps routine package checks practical while preserving full analysis code.

# 1. Learning objectives

You will learn to:

- inspect available prior classes;
- specify scientifically meaningful priors;
- perform prior predictive checks;
- fit a Bayesian multilevel model;
- inspect posterior draws;
- understand credible intervals and direct probability statements;
- use posterior predictive checks;
- use leave-one-out predictive assessment;
- distinguish Bayesian evidence from frequentist p-values.

# 2. Data

```{r data}
library(mixedFlowR)
z <- mixed_data("longitudinal")
head(z)
```

# 3. Inspect prior structure

`mixed_prior()` delegates to the `brms` prior-inspection machinery.

```{r prior-classes}
if (requireNamespace("brms", quietly = TRUE)) {
  prior_classes <- mixed_prior(
    height ~ treatment * time + (time | subject),
    data = z
  )

  prior_classes
}
```

This step is light enough to be evaluated when `brms` is installed because it does not sample the posterior.

# 4. Why priors matter

A prior should be interpreted on the scale of its parameter.

For example:

- intercept prior: plausible plant height at the reference treatment/time;
- time slope prior: plausible rate of change per time unit;
- group-level SD prior: plausible between-subject heterogeneity;
- residual SD prior: plausible observation-level variability.

Avoid choosing a prior only because it is described generically as "weakly informative." Translate it into the units of the agronomic response.

# 5. Example prior specification

```{r explicit-priors, eval=FALSE}
priors <- c(
  brms::prior(normal(20, 10), class = "Intercept"),
  brms::prior(normal(0, 3), class = "b"),
  brms::prior(exponential(1), class = "sd"),
  brms::prior(exponential(1), class = "sigma")
)
```

These are teaching values, not universal priors. Replace them with values justified by measurement scale and domain knowledge.

# 6. Prior predictive simulation

```{r prior-predictive, eval=FALSE}
prior_fit <- mixed_prior_predictive(
  height ~ treatment * time + (time | subject),
  data   = z,
  prior  = priors,
  chains = 2,
  iter   = 1000,
  seed   = 260818
)
```

A prior predictive check asks whether the prior model generates plausible data before the observed response informs the posterior.

If simulated heights are physically absurd, revise the priors before fitting the posterior.

# 7. Fit the Bayesian multilevel model

```{r bayes-fit, eval=FALSE}
bfit <- mixed_bayes(
  height ~ treatment * time + (time | subject),
  data   = z,
  prior  = priors,
  chains = 4,
  iter   = 4000,
  seed   = 260818
)
```

For local exploratory runs, a smaller number of iterations may be used to verify syntax, but scientific inference should use adequate effective sample size and convergence diagnostics.

# 8. Posterior draws

```{r posterior-draws, eval=FALSE}
draws <- mixed_posterior(bfit)
draws
```

Select variables:

```{r posterior-selected, eval=FALSE}
time_draws <- mixed_posterior(
  bfit,
  variables = "b_time"
)
```

Posterior draws allow direct calculation of derived scientific quantities.

# 9. Credible intervals

A 95% posterior credible interval contains 95% of posterior probability for the parameter under the fitted Bayesian model.

It should not be described as "the same as a confidence interval but Bayesian."

The interpretation depends on:

- likelihood;
- prior;
- hierarchical structure;
- posterior computation.

# 10. Direct probability statements

With posterior draws, questions such as these are natural:

- `Pr(treatment effect > 0 | data, model)`;
- `Pr(yield advantage > agronomic threshold | data, model)`;
- `Pr(random-slope SD exceeds a meaningful value | data, model)`.

A biologically meaningful threshold is often more informative than `Pr(beta > 0)` alone.

```{r posterior-probability, eval=FALSE}
d <- mixed_posterior(bfit, variables = "b_treatmentT2")
mean(d$b_treatmentT2 > 0)
mean(d$b_treatmentT2 > 2)
```

# 11. Posterior predictive checks

```{r ppc-density, eval=FALSE}
mixed_pp_check(
  bfit,
  type = "dens_overlay",
  ndraws = 100
)
```

Alternative ECDF check:

```{r ppc-ecdf, eval=FALSE}
mixed_pp_check(
  bfit,
  type = "ecdf_overlay",
  ndraws = 100
)
```

A model can have apparently precise posterior coefficients and still fail to reproduce key features of the observed data.

# 12. Convergence diagnostics

Inspect at least:

- R-hat;
- effective sample size;
- divergent transitions;
- energy or treedepth warnings when relevant;
- trace plots for consequential parameters.

Do not interpret posterior summaries if the sampler has not adequately explored the posterior.

# 13. LOO predictive assessment

```{r loo-single, eval=FALSE}
loo_fit <- mixed_loo(bfit)
loo_fit
```

Compare two models:

```{r loo-compare, eval=FALSE}
b0 <- mixed_bayes(
  height ~ time + (time | subject),
  data = z,
  prior = priors,
  chains = 4,
  iter = 4000,
  seed = 260818
)

mixed_loo(b0, bfit)
```

LOO comparison is predictive assessment, not a universal proof of the scientific truth of one model.

# 14. Bayesian binomial example

The disease-incidence dataset can be modeled with a binomial multilevel model.

```{r bayes-binomial, eval=FALSE}
b <- mixed_data("binomial")

b_bin <- mixed_bayes(
  diseased | trials(total) ~ treatment + (1 | block),
  data = b,
  family = binomial(),
  chains = 4,
  iter = 4000,
  seed = 260818
)
```

This illustrates that the Bayesian workflow extends naturally to non-Gaussian multilevel outcomes.

# 15. Bayesian distributional model

`brms` can model distributional parameters as predictors. That capability overlaps conceptually with GAMLSS but uses a Bayesian framework.

A generic template:

```{r distributional, eval=FALSE}
bf_dist <- brms::bf(
  height ~ treatment * time + (1 | subject),
  sigma ~ treatment
)

b_dist <- mixed_bayes(
  bf_dist,
  data = z,
  prior = distributional_priors,  # define scientifically justified priors before fitting
  chains = 4,
  iter = 4000,
  seed = 260818
)
```

`distributional_priors` is deliberately left as a user-defined object. Priors for distributional parameters should be chosen for the actual parameterization rather than copied blindly.

# 16. Sensitivity to priors

A Bayesian analysis should examine whether consequential conclusions change under reasonable prior alternatives.

A useful sensitivity design is:

1. weakly informative prior;
2. domain-informed prior;
3. more skeptical prior around a scientifically important treatment effect.

If conclusions change dramatically, report that prior sensitivity rather than hiding it.

# 17. Bayesian versus frequentist mixed models

These approaches can use similar hierarchical likelihood structures but differ in inference.

| Question | Frequentist LMM | Bayesian multilevel |
|---|---|---|
| fixed-effect uncertainty | sampling-based CI/test | posterior distribution |
| random effects | estimated covariance + BLUPs | posterior distributions |
| prior information | not explicit prior | explicit prior model |
| model checking | residual diagnostics | residual + posterior predictive checks |
| probability parameter exceeds threshold | indirect | direct posterior calculation |

Neither framework removes the need to respect experimental design.

# 18. Computational recommendations

For reproducible scientific runs:

- use a documented backend;
- save the seed;
- record `sessionInfo()`;
- inspect sampler diagnostics;
- use sufficient chains and iterations;
- avoid recompiling unchanged models unnecessarily in teaching loops;
- consider `cmdstanr` when appropriate and supported locally.

# 19. Reporting checklist

Report:

- likelihood/family;
- fixed and group-level structure;
- prior distributions and rationale;
- chain count, iterations, warmup if relevant;
- sampler/backend;
- R-hat and effective sample size;
- posterior interval type;
- posterior predictive checks;
- LOO or other predictive assessment if used;
- prior sensitivity for important conclusions.

# 20. Exercises

1. Propose an intercept prior for a crop height measured in centimeters and justify it physically.
2. Create a skeptical prior for a treatment effect larger than 10 cm.
3. Explain why a posterior predictive check can fail even if every R-hat is close to 1.
4. Compare the interpretation of a 95% credible interval with a 95% confidence interval.

# 21. Key take-home messages

- Priors are part of the model and should be transparent.
- Prior predictive checking should precede posterior fitting when priors matter.
- Posterior intervals support direct probability statements under the fitted model.
- Sampler convergence is a prerequisite for interpretation.
- Posterior predictive checks evaluate whether the model can reproduce relevant data features.
- Bayesian complexity should answer a scientific question, not serve as decoration.

# 22. Next vignette

Continue with `11-nonlinear-random-regression.Rmd` for logistic/Gompertz growth, dose response, generic nonlinear mixed models, and random regression.

# References

The `brms` framework and its multilevel Bayesian formulation are described by Buerkner [@burkner2017].
