---
title: "Diagnostics and Model Comparison"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Diagnostics and Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

Model comparison should follow model diagnosis, not replace it. This vignette develops a structured workflow for convergence, singularity, residual patterns, simulation diagnostics, influence, likelihood-based comparisons, small-sample tests, and parametric-bootstrap comparisons.

# 1. Learning objectives

You will learn to:

- interpret convergence and singularity diagnostics;
- inspect residual data before inference;
- use `DHARMa` simulation diagnostics for supported GLMMs;
- conduct grouped influence analysis;
- compare candidate models using information criteria without automatic selection;
- perform fixed-effect LRTs under ML;
- use Kenward-Roger and parametric-bootstrap comparisons for suitable LMMs;
- understand why a valid comparison depends on the estimand and fitting method.

# 2. Diagnose a Gaussian LMM

```{r data}
library(mixedFlowR)
z <- mixed_data("longitudinal")
```

```{r lmm}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit <- mixed_fit(
    height ~ treatment * time + (time | subject),
    data = z
  )
}
```

```{r diagnose}
if (exists("fit")) {
  diag <- mixed_diagnose(fit)
  diag
}
```

The diagnostic object contains convergence status, messages, singularity where supported, residual data, variance summaries, and optional simulation residuals.

# 3. Residual plot

```{r residual-plot}
if (exists("fit")) {
  mixed_plot(fit, type = "residuals")
}
```

Look for:

- nonlinear residual trend;
- funnel-shaped variance;
- clusters of extreme residuals;
- unexplained grouping pattern;
- serial or spatial dependence that requires a dedicated test or plot.

# 4. Singularity

A singular LMM has one or more random-effect dimensions estimated on or near a boundary. Common signs include a variance near zero or an extreme random-effect correlation.

A singular fit can mean:

- the data do not support the requested random structure;
- the design provides too little repeated information;
- a random slope is unnecessary;
- predictors are poorly scaled or nearly redundant.

It does **not** mean that the optimizer necessarily failed.

# 5. Create a deliberately demanding random structure

```{r demanding-fit}
if (requireNamespace("lme4", quietly = TRUE)) {
  z_sparse <- subset(z, time %in% c(0, 1, 5))
  demanding <- mixed_random_slopes(
    z_sparse,
    response = "height",
    fixed = "treatment * time",
    slope = "time",
    group = "subject",
    correlated = TRUE
  )
  mixed_diagnose(demanding)
}
```

This example is designed to encourage diagnosis, not to guarantee a singular fit on every backend version.

# 6. Simulation diagnostics for a count GLMM

```{r count-model}
cnt <- mixed_data("counts")

if (requireNamespace("glmmTMB", quietly = TRUE)) {
  nb <- mixed_count(
    count ~ treatment * time + (1 | plot),
    data = cnt,
    family = "nbinom2"
  )
}
```

```{r count-dharma}
if (
  exists("nb") &&
  requireNamespace("DHARMa", quietly = TRUE)
) {
  nb_diag <- mixed_diagnose(nb, simulate = TRUE, nsim = 500)
  nb_diag$simulation
}
```

Simulation residuals are particularly helpful for detecting overdispersion, zero inflation, and nonuniform residual distributions in discrete models.

# 7. Influence by cluster

For `lme4` models, grouped deletion diagnostics can be requested through `influence.ME`.

```{r influence}
if (
  exists("fit") &&
  requireNamespace("influence.ME", quietly = TRUE)
) {
  infl <- mixed_influence(
    fit,
    group = "subject"
  )

  infl
}
```

Influence means that removing a unit changes an estimate or inferential quantity materially. It is not the same as proving a data error.

# 8. Information criteria as descriptors

Fit two models under ML because their fixed-effect structures differ.

```{r ml-models}
if (requireNamespace("lme4", quietly = TRUE)) {
  m0_ml <- mixed_fit(
    height ~ time + (1 | subject),
    data = z,
    REML = FALSE
  )

  m1_ml <- mixed_fit(
    height ~ treatment + time + (1 | subject),
    data = z,
    REML = FALSE
  )
}
```

```{r information}
if (exists("m0_ml") && exists("m1_ml")) {
  mixed_compare(
    no_treatment = m0_ml,
    treatment    = m1_ml,
    method       = "information"
  )
}
```

AIC and BIC provide penalized likelihood descriptors. They do not know the experimental design or biological importance.

# 9. Likelihood-ratio test under ML

```{r lrt}
if (exists("m0_ml") && exists("m1_ml")) {
  mixed_compare(
    m0_ml,
    m1_ml,
    method = "LRT"
  )
}
```

When fixed effects differ, the package blocks this LRT if an `lmer` model is still fitted by REML.

# 10. Deliberate invalid comparison

```{r invalid-reml-lrt, eval=FALSE}
m0_reml <- mixed_fit(height ~ time + (1 | subject), z, REML = TRUE)
m1_reml <- mixed_fit(height ~ treatment + time + (1 | subject), z, REML = TRUE)
mixed_compare(m0_reml, m1_reml, method = "LRT")
```

This chunk is deliberately not evaluated because it is intended to show a safeguarded error.

# 11. Kenward-Roger comparison

Fit two nested LMMs under the conventional REML setting.

```{r kr-models}
if (requireNamespace("lme4", quietly = TRUE)) {
  kr_small <- mixed_fit(
    height ~ time + (1 | subject),
    data = z
  )

  kr_large <- mixed_fit(
    height ~ treatment + time + (1 | subject),
    data = z
  )
}
```

```{r kr}
if (
  exists("kr_small") && exists("kr_large") &&
  requireNamespace("pbkrtest", quietly = TRUE)
) {
  mixed_compare(
    kr_small,
    kr_large,
    method = "Kenward-Roger"
  )
}
```

Kenward-Roger adjusts the small-sample reference distribution and covariance estimation for fixed-effect testing in appropriate Gaussian LMM settings.

# 12. Parametric-bootstrap comparison

Use a small simulation count for vignette execution.

```{r pb-demo}
if (
  exists("kr_small") && exists("kr_large") &&
  requireNamespace("pbkrtest", quietly = TRUE)
) {
  pb_demo <- mixed_compare(
    kr_small,
    kr_large,
    method = "parametric-bootstrap",
    nsim = 99,
    seed = 260818
  )

  pb_demo
}
```

Production-scale example:

```{r pb-production, eval=FALSE}
pb_full <- mixed_compare(
  kr_small,
  kr_large,
  method = "parametric-bootstrap",
  nsim = 4999,
  seed = 260818
)
```

The production chunk is not evaluated during vignette build because thousands of refits are intentionally expensive.

# 13. Compare covariance candidates separately

```{r covariance-models}
if (requireNamespace("nlme", quietly = TRUE)) {
  ar1 <- mixed_ar1(z, "height", "treatment*time", "time", "subject")
  car1 <- mixed_car1(z, "height", "treatment*time", "time", "subject")
}
```

```{r covariance-compare}
if (exists("ar1") && exists("car1")) {
  mixed_covariance_compare(
    AR1 = ar1,
    CAR1 = car1
  )
}
```

The fixed mean model should be kept comparable while evaluating covariance structure.

# 14. A model-comparison decision tree

Use the following logic.

**Question A: Are fixed effects different?**

- For `lmer` likelihood-ratio comparison, fit under ML.
- For small-sample Gaussian LMM inference, consider Kenward-Roger or parametric bootstrap when appropriate.

**Question B: Are random or covariance structures different?**

- Keep the scientific fixed-effect estimand comparable.
- Use likelihood descriptors and diagnostics.
- Prefer structures supported by the design and repeated-measure geometry.

**Question C: Are model families different?**

- Diagnose distributional adequacy.
- Compare predictive implications and biological process assumptions.
- Do not rely on a single criterion.

# 15. Diagnostic table output

```{r diagnostics-table}
if (exists("diag")) {
  mixed_table(
    diag,
    component = "diagnostics",
    format = "markdown",
    caption = "Mixed-model diagnostic summary"
  )
}
```

# 16. What not to do

Avoid workflows such as:

1. generate ten random structures;
2. rank by AIC;
3. keep the minimum;
4. ignore convergence warnings;
5. report only the final p-values.

This process disconnects modeling from experimental design and understates selection uncertainty.

# 17. Practice exercises

1. Fit random-intercept and random-slope models with the same fixed effects. Compare their diagnostics before AIC.
2. Fit Poisson and negative-binomial count models. Ask whether the more flexible distribution solves a residual problem.
3. Compare AR(1) and heterogeneous AR(1) models. State the additional scientific assumption introduced by heterogeneous residual variances.
4. Refit a fixed-effect comparison under REML and verify that the guarded LRT refuses the invalid comparison.

# 18. Reporting checklist

For every model comparison, report:

- models being compared;
- what differs between them;
- fitting method;
- comparison method;
- diagnostic status;
- rationale for the preferred representation;
- whether substantive conclusions are stable.

# 19. Key take-home messages

- Diagnostics precede model ranking.
- Convergence and singularity are different concepts.
- Information criteria are descriptors, not scientific decision rules.
- Fixed-effect LRTs with `lmer` require appropriate ML fitting when the fixed structure changes.
- Influence identifies sensitivity, not automatically erroneous observations.
- Small-sample methods answer specific inferential questions and do not replace design reasoning.

# 20. Next vignette

Continue with `06-robust-sensitivity.Rmd` for robust estimation, contamination screens, and cluster-robust covariance.
