---
title: "Teaching Mixed Models with mixedFlowR Tours"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Teaching Mixed Models with mixedFlowR Tours}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

`mixedFlowR` was designed not only as an analysis interface but also as a teaching scaffold. The `mixed_tour()` function exposes conceptual sequences that can be used in lectures, workshops, laboratory exercises, self-study, and graduate methods courses.

This vignette turns those tours into a ten-session instructional pathway.

# 1. Learning objectives

After completing this vignette, an instructor or self-learner should be able to:

- use `mixed_tour()` to retrieve teaching sequences;
- organize topics from design to advanced mixed models;
- separate conceptual learning from computational complexity;
- build classroom exercises around frozen teaching datasets;
- integrate figures, tables, diagnostics, and interpretation;
- use the same dataset to show progressive model refinement.

# 2. Available tours

```{r tours}
library(mixedFlowR)

topics <- c(
  "splitplot",
  "random_effects",
  "covariance",
  "robust",
  "gamlss",
  "bayes",
  "spatial",
  "met"
)

lapply(topics, mixed_tour)
```

# 3. Split-plot tour

```{r splitplot-tour}
tour_split <- mixed_tour("splitplot")
tour_split$steps
```

Run the lightweight artifact.

```{r splitplot-run}
tour_split_run <- mixed_tour("splitplot", run = TRUE)
tour_split_run$artifact
```

The sequence moves from experimental units to error strata and then to a mixed representation.

## Classroom prompt

Ask students to identify which error term would be wrong if every subplot were treated as an independent variety replicate.

# 4. Random-effects tour

```{r random-tour}
mixed_tour("random_effects")$steps
```

Suggested demonstration:

```{r random-demo}
z <- mixed_data("longitudinal")

if (requireNamespace("lme4", quietly = TRUE)) {
  ri <- mixed_random_intercepts(z, "height", "treatment*time", "subject")
  rs <- mixed_random_slopes(z, "height", "treatment*time", "time", "subject")

  mixed_plot(rs, type = "random")
}
```

## Classroom prompt

What changes when a random slope is added? Which parameter represents subject-to-subject heterogeneity in growth rate?

# 5. Covariance tour

```{r covariance-tour}
mixed_tour("covariance")$steps
```

Demonstrate AR(1) and heterogeneity.

```{r covariance-demo}
if (requireNamespace("nlme", quietly = TRUE)) {
  ar1 <- mixed_ar1(z, "height", "treatment*time", "time", "subject")
  het <- mixed_heterogeneity(
    height ~ treatment * time,
    z,
    random = ~ 1 | subject,
    variance_group = "treatment"
  )
}
```

## Classroom prompt

Explain why random intercept covariance and residual AR(1) covariance are not the same source of dependence.

# 6. Robust tour

```{r robust-tour}
mixed_tour("robust")$steps
```

Create a teaching contamination scenario.

```{r robust-demo}
z_cont <- z
z_cont$height[1] <- z_cont$height[1] + 30

if (
  requireNamespace("lme4", quietly = TRUE) &&
  requireNamespace("robustlmm", quietly = TRUE)
) {
  classical <- mixed_fit(height ~ treatment*time + (1|subject), z_cont)
  robust <- mixed_robust(height ~ treatment*time + (1|subject), z_cont)
  mixed_compare_robust(classical, robust)
}
```

## Classroom prompt

A large classical-versus-robust shift is observed. Is deleting the flagged observation justified? What information is needed before exclusion?

# 7. GAMLSS tour

```{r gamlss-tour}
mixed_tour("gamlss")$steps
```

```{r gamlss-demo}
if (requireNamespace("gamlss", quietly = TRUE)) {
  g_scale <- mixed_gamlss(
    height ~ treatment*time,
    z,
    family = "NO",
    sigma.formula = ~ treatment,
    random = "subject"
  )

  head(mixed_gamlss_parameters(g_scale, c("mu", "sigma")))
}
```

## Classroom prompt

What biological question is asked by putting treatment in `sigma.formula`? How is it different from asking whether treatment changes mean height?

# 8. Bayesian tour

```{r bayes-tour}
mixed_tour("bayes")$steps
```

Prior inspection can be shown without sampling.

```{r bayes-prior}
if (requireNamespace("brms", quietly = TRUE)) {
  mixed_prior(
    height ~ treatment*time + (time|subject),
    data = z
  )
}
```

Full posterior fitting should be demonstrated in a prepared environment rather than forced during every package check.

```{r bayes-full, eval=FALSE}
bfit <- mixed_bayes(
  height ~ treatment*time + (time|subject),
  z,
  prior = priors,
  chains = 4,
  iter = 4000,
  seed = 260818
)

mixed_pp_check(bfit)
```

## Classroom prompt

What does a posterior predictive check assess that R-hat does not?

# 9. Spatial tour

```{r spatial-tour}
mixed_tour("spatial")$steps
```

```{r spatial-demo}
s <- mixed_data("spatial")

if (requireNamespace("SpATS", quietly = TRUE)) {
  sp <- mixed_spatial_field(s, "yield", "genotype", "col", "row", random = ~ block)
  mixed_plot(sp, type = "spatial")
}
```

## Classroom prompt

Which pattern in the spatial surface could reflect soil heterogeneity? Why must treatment layout be inspected before attributing the pattern to field trend?

# 10. MET tour

```{r met-tour}
mixed_tour("met")$steps
```

```{r met-demo}
m <- mixed_data("met")

if (requireNamespace("sommer", quietly = TRUE)) {
  gxe <- mixed_gxe(m, "yield", "genotype", "environment", "fa", nPC = 2)
}
```

## Classroom prompt

Why should an FA factor not automatically be named "rainfall" or "temperature"?

# 11. A ten-session course design

| Session | Core question | Main functions |
|---|---|---|
| 1 | What is the experimental unit? | `mixed_design_audit()`, `mixed_hierarchy()` |
| 2 | Why does split-plot have two errors? | `fit_legacy_splitplot()`, `splitplot_ems()` |
| 3 | What is partial pooling? | `mixed_random_intercepts()`, `mixed_blup()` |
| 4 | Why random slopes? | `mixed_random_slopes()`, `mixed_covariance_plot()` |
| 5 | How do non-Gaussian responses change the model? | `mixed_count()`, `mixed_binomial()`, `mixed_beta()` |
| 6 | Where does repeated-measure correlation enter? | `mixed_ar1()`, `mixed_heterogeneity()` |
| 7 | How should dose be interpreted? | `mixed_trend()`, `mixed_curve()`, `mixed_optimum()` |
| 8 | How sensitive are conclusions? | `mixed_robust()`, `mixed_boot()` |
| 9 | Why model distributions/spatial/MET structure? | `mixed_gamlss()`, `mixed_spatial_field()`, `mixed_gxe()` |
| 10 | How do we communicate and reproduce the analysis? | `mixed_plot()`, `mixed_table()`, `mixed_report()` |

# 12. Three-level teaching pattern

For every major function, use three examples.

**Level 1: minimal syntax**  
Show what the function does.

**Level 2: realistic agronomic problem**  
Connect design, data, and interpretation.

**Level 3: problematic or discrepant scenario**  
Show what can go wrong, such as missing cells, singularity, contamination, overdispersion, boundary values, or spatial confounding.

This pattern is consistent with the public API example catalog.

# 13. Figures as teaching objects

Do not show a figure only as decoration. Ask students to identify:

- observed versus fitted information;
- uncertainty;
- grouping structure;
- residual pattern;
- extrapolation;
- scale and units.

# 14. Tables as teaching objects

A table exercise should ask:

- What is the estimand?
- What does the standard error represent?
- Which random component is associated with which experimental unit?
- What adjustment was used for multiple comparisons?
- Is the estimate on the link scale or response scale?

# 15. Suggested practical assessment

Give students one frozen dataset and require:

1. design audit;
2. raw-data plot;
3. justified model specification;
4. diagnostics;
5. one alternative/sensitivity analysis;
6. one publication-ready figure;
7. one publication-ready table;
8. a 200-word interpretation;
9. reproducibility information.

Grade the reasoning, not only whether the software runs.

# 16. Teaching Bayesian and heavy modules responsibly

For computationally expensive modules:

- run prevalidated examples in class;
- provide the full code;
- use `eval=FALSE` in routine package vignette builds;
- give students smaller syntax-check configurations;
- separate computational convenience from final analysis settings.

# 17. Self-study checklist

A learner is ready to advance when they can explain:

- the experimental unit without looking at the formula;
- why a random effect was included;
- why a response family was selected;
- why a covariance structure is plausible;
- what a diagnostic failure means;
- what an EMM, trend, BLUP, and posterior draw represent;
- what uncertainty measure is shown in a figure.

# 18. Exercises

1. Use `mixed_tour()` to write a five-slide lesson outline for random effects.
2. Create a contamination scenario and ask a peer to decide whether exclusion is justified.
3. Compare a qualitative treatment analysis with a quantitative dose analysis of the same data.
4. Build a one-page practical report containing one figure, one table, and four interpretive bullets.

# 19. Key take-home messages

- Teaching should preserve the connection between experimental design and model structure.
- The same dataset can illustrate progressive model refinement.
- Heavy computation should not force all vignettes to `eval=FALSE`.
- Interpretation exercises are as important as code exercises.
- A good mixed-model course teaches what **not** to model as well as what can be modeled.

# 20. What comes next

Stage 5 closes the tutorial sequence with the state of the art, reproducibility/validation, a complete agronomic workflow, and the full API example catalog.
