---
title: "Validation and Reproducibility"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Validation and Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

A statistical package is not validated because its examples look plausible. Validation requires agreement with backend engines, known-design behavior, deterministic datasets and seeds, diagnostics, automated tests, dependency checks, and a reproducible local check workflow.

This vignette explains how to validate `mixedFlowR` without confusing documentation examples with evidence of numerical correctness.

# 1. Learning objectives

You will learn to:

- distinguish static validation from runtime statistical validation;
- inspect package capabilities and block coverage;
- verify frozen teaching datasets;
- reproduce seeds and simulation scenarios;
- run unit and golden tests locally;
- regenerate `roxygen2` documentation and `NAMESPACE`;
- run `R CMD check` through `rcmdcheck`;
- validate optional backends separately;
- document a validation session.

# 2. Validation layers

A useful validation hierarchy is:

1. **static integrity**: files, exports, examples, placeholders, manifests;
2. **unit tests**: functions return expected classes and safeguards;
3. **golden backend tests**: wrappers agree with direct backend fits;
4. **design benchmarks**: legacy and mixed representations align when theory says they should;
5. **simulation validation**: known generating parameters are recovered with appropriate coverage;
6. **cross-platform checks**: Windows, Linux, and supported R versions;
7. **scientific review**: methods and interpretations remain defensible.

# 3. Inspect capabilities

```{r capabilities}
library(mixedFlowR)
mixed_capabilities()
```

This identifies which optional validation paths can run on the current computer.

# 4. Confirm all 70 blocks are registered

```{r blocks}
blocks <- mixed_blocks()
blocks
nrow(blocks)
```

The block registry is a coverage map. It does not prove that every algorithm is numerically correct.

# 5. Frozen teaching datasets

```{r datasets}
names_to_check <- c(
  "splitplot", "longitudinal", "counts", "binomial", "beta",
  "spatial", "met", "growth", "dose_response"
)

lapply(names_to_check, function(x) dim(mixed_data(x)))
```

Frozen teaching data allow documentation and tests to refer to stable inputs. They should not be regenerated silently during routine checks.

# 6. Reproducible seeds

```{r seed}
set.seed(260818)
rnorm(5)
set.seed(260818)
rnorm(5)
```

A seed helps reproduce stochastic computation. It does not make a stochastic estimator deterministic across every R/backend/platform version if underlying numerical algorithms change.

# 7. Golden test principle

A wrapper around `lme4` should agree with direct `lme4` fitting when both represent the same model.

Example test logic:

```{r golden-example, eval=FALSE}
d <- mixed_data("longitudinal")

wrapped <- mixed_fit(
  height ~ treatment*time + (1|subject),
  d,
  engine = "lme4"
)

direct <- lme4::lmer(
  height ~ treatment*time + (1|subject),
  data = d
)

testthat::expect_equal(
  lme4::fixef(wrapped$model),
  lme4::fixef(direct),
  tolerance = 1e-8
)
```

Tests should compare quantities that are theoretically expected to agree, not merely visually similar output.

# 8. Legacy split-plot benchmark

Balanced split-plot data provide an important design benchmark.

```{r legacy-benchmark}
d <- mixed_data("splitplot")
d$nitrogen_f <- factor(d$nitrogen)

legacy <- fit_legacy_splitplot(
  d, "yield", "block", "variety", "nitrogen_f"
)
splitplot_anova(legacy)
```

```{r legacy-mixed-benchmark}
if (requireNamespace("lme4", quietly = TRUE)) {
  compare_legacy_mixed(
    d, "yield", "block", "variety", "nitrogen_f"
  )
}
```

The benchmark should test equivalence properties that are expected in balanced settings and document where inferential conventions differ.

# 9. Safeguard tests are scientific tests

Examples include verifying that:

- beta models reject 0/1 values when the selected family requires `(0,1)`;
- LRTs of different fixed-effect structures under REML are blocked;
- missing optional backends produce informative errors;
- bootstrap seeds are recorded;
- contamination flags never delete observations automatically.

These are not cosmetic checks. They protect scientific interpretation.

# 10. Run the package tests

From the package root:

```{r testthat, eval=FALSE}
testthat::test_local()
```

For more diagnostic output:

```{r testthat-reporter, eval=FALSE}
testthat::test_local(reporter = "progress")
```

# 11. Regenerate roxygen2 documentation and NAMESPACE

The authoritative export/import source is the roxygen documentation in `R/`.

```{r roxygen, eval=FALSE}
roxygen2::roxygenise()
```

Then inspect changes to:

- `NAMESPACE`;
- `man/*.Rd`;
- any unexpectedly removed exports.

Do not hand-maintain a generated `NAMESPACE` in parallel with roxygen tags.

# 12. Load the source tree without installing

```{r load-all, eval=FALSE}
pkgload::load_all()
```

This is useful for development iteration. It does not replace installation and `R CMD check`.

# 13. Run R CMD check through rcmdcheck

```{r rcmdcheck, eval=FALSE}
rcmdcheck::rcmdcheck(
  args = c("--as-cran"),
  error_on = "warning"
)
```

Investigate every warning or note. Do not suppress a diagnostic merely to obtain a green badge.

# 14. Vignette evaluation policy

Most tutorial chunks use:

```r
knitr::opts_chunk$set(eval = TRUE)
```

with optional backends guarded by `requireNamespace()`.

Heavy chunks are intentionally `eval=FALSE`, including:

- thousands of bootstrap refits;
- production-scale Bayesian sampling;
- deliberately failing safeguard examples;
- examples that write persistent files during package build.

This policy keeps the documentation executable without making routine checks impractical.

# 15. Validate optional backends individually

A local validation matrix should include at least:

| Backend | Validation target |
|---|---|
| `lme4` | LMM/GLMM agreement and random structures |
| `nlme` | AR1/CAR1/heterogeneity/NLMM |
| `glmmTMB` | count, beta, ZI, Toeplitz, spatial covariance |
| `robustlmm` | robust estimate agreement |
| `confintROB` | BCa/wild interval routes |
| `emmeans` | EMM/trend consistency |
| `gamlss` | parameter extraction/quantiles |
| `brms` | prior, posterior, PPC, LOO |
| `SpATS` | field model and surface extraction |
| `sommer` | MET and GxE covariance |

# 16. Heavy validation should be separated from CRAN-style smoke tests

A simulation study for interval coverage or GxE recovery may take minutes or hours. Keep it in a dedicated validation script rather than forcing it into every unit-test run.

A reproducible heavy-analysis protocol should record:

- scenario ID;
- seed;
- generating parameters;
- number of simulations;
- failed fit count;
- convergence count;
- bias;
- RMSE;
- interval coverage;
- computing environment.

# 17. Simulation validation concept

For a parameter `theta`, assess:

- bias: average estimate minus true value;
- RMSE: root mean squared estimation error;
- coverage: fraction of intervals containing the true value;
- convergence/failure rate;
- sensitivity to sample size and design imbalance.

Do not tune methods using only rejection rates or p-values.

# 18. Reproducibility metadata

```{r session-info}
utils::sessionInfo()
```

Also record:

```{r package-versions}
pkgs <- c("mixedFlowR", "lme4", "nlme", "glmmTMB", "emmeans")
sapply(pkgs, function(p) {
  if (requireNamespace(p, quietly = TRUE)) as.character(utils::packageVersion(p)) else NA_character_
})
```

# 19. Hashes and frozen artifacts

For archived validation, calculate file hashes for:

- frozen datasets;
- simulation inputs;
- release tarball;
- reference metadata;
- key expected-output files if used as golden artifacts.

A hash detects unintended file change; it does not establish scientific validity.

# 20. Cross-platform validation

At minimum, validate on:

- Windows with the intended user-facing R version;
- a Linux CI environment;
- the oldest supported R version where practical;
- a current R-devel check for future compatibility.

Optional compiled backends may expose platform-specific issues that are invisible in static validation.

# 21. A local release gate

Before release, require:

1. `roxygen2::roxygenise()` clean;
2. `testthat` clean;
3. static audit clean;
4. heavy simulation battery reviewed;
5. `R CMD check --as-cran` without unexplained warnings;
6. all 20 vignettes render with expected optional-package behavior;
7. README examples consistent with exported API;
8. source archive inspected after extraction;
9. reference metadata reviewed;
10. package version and NEWS updated.

# 22. What this vignette does not claim

The presence of code in a vignette is not runtime validation evidence. The actual runtime environment must execute the relevant tests.

Likewise, passing `R CMD check` means the package satisfies important technical checks; it does not prove statistical optimality for every scientific application.

# 23. Exercises

1. Run `mixed_capabilities()` on two computers and compare optional backend availability.
2. Create an invalid beta value and write a test expecting an error.
3. Write a golden test comparing `mixed_fit()` with direct `lme4::lmer()`.
4. Create a simulation scenario with a known random-intercept variance and assess recovery.
5. Render a vignette after temporarily removing one optional package and confirm graceful degradation.

# 24. Key take-home messages

- Static validation and runtime validation are different layers.
- Wrappers should be tested against their authoritative backends.
- Design benchmarks are as important as software benchmarks.
- Seeds, versions, and frozen inputs are part of reproducibility.
- Heavy simulation belongs in a separate validation battery.
- `R CMD check` is necessary but not sufficient for scientific validation.

# 25. Next vignette

Continue with `19-complete-agronomic-workflow.Rmd` for an end-to-end analysis that combines design, modeling, diagnosis, inference, sensitivity, visualization, and reporting.
