---
title: "Complete Agronomic Workflow: Split-Plot Nitrogen Response"
author: "mixedFlowR tutorial series"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Complete Agronomic Workflow: Split-Plot Nitrogen Response}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE, error = FALSE,
  collapse = TRUE, comment = "#>", fig.align = "center",
  fig.width = 7, fig.height = 5, dpi = 150
)
set.seed(260818)
```


# Purpose

This capstone vignette integrates the major concepts of the tutorial series in one coherent agronomic analysis. The example is a randomized complete-block split-plot trial with variety in whole plots, nitrogen dose in subplots, and yield as the response.

The purpose is not to use every advanced model. The purpose is to show **when** each extension becomes relevant and when the simpler analysis should be retained.

# 1. Learning objectives

You will learn to execute a complete workflow:

1. define the scientific question;
2. audit the randomization and experimental units;
3. explore raw data;
4. use legacy ANOVA as a design check;
5. fit the design-correct LMM;
6. diagnose the fit;
7. interpret qualitative and quantitative effects;
8. evaluate uncertainty;
9. conduct robust and bootstrap sensitivity analyses;
10. produce publication-ready figures and tables;
11. generate a reproducibility record;
12. decide whether advanced GAMLSS/Bayesian/spatial extensions are justified.

# 2. Scientific question

Assume the study asks:

> Do varieties differ in grain yield, how does yield respond to nitrogen dose, and does the nitrogen response depend on variety when the split-plot randomization is represented correctly?

Secondary questions are:

- Are conclusions sensitive to influential observations?
- Are uncertainty intervals stable under bootstrap resampling?
- Is there evidence that a more complex distributional or covariance model is needed?

# 3. Load data and record capabilities

```{r data}
library(mixedFlowR)
d <- mixed_data("splitplot")
head(d)
```

```{r capabilities}
mixed_capabilities()
```

# 4. Define the estimands before fitting

**Variety:** qualitative whole-plot treatment.  
**Nitrogen:** quantitative subplot treatment.  
**Variety × nitrogen:** modification of dose response by variety.  
**Block and block-by-variety:** randomization-induced grouping components.

This declaration prevents the analysis from drifting into arbitrary pairwise tests or automatic model selection.

# 5. Audit the design

```{r audit}
audit <- mixed_design_audit(
  d,
  response = "yield",
  block = "block",
  whole = "variety",
  sub = "nitrogen"
)
audit
```

```{r hierarchy}
mixed_hierarchy(d, block = "block", whole = "variety", sub = "nitrogen")
```

```{r balance}
mixed_balance(d, c("block", "variety", "nitrogen"))
```

# 6. Explicitly identify whole plots

```{r whole-plots}
d$whole_plot <- interaction(d$block, d$variety, drop = TRUE)

mixed_pseudorep(
  d,
  response = "yield",
  cluster = "whole_plot",
  treatment = "variety"
)
```

The variety treatment is replicated at the whole-plot level, not at the subplot-observation level.

# 7. Raw-data exploration

```{r raw-plot, fig.height=5.2}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(
    d,
    ggplot2::aes(nitrogen, yield, colour = variety)
  ) +
    ggplot2::geom_point(alpha = 0.7) +
    ggplot2::geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
    ggplot2::facet_wrap(~ block) +
    ggplot2::labs(
      x = "Nitrogen dose",
      y = "Yield",
      colour = "Variety",
      title = "Observed split-plot response by block"
    ) +
    ggplot2::theme_minimal(base_size = 11)
}
```

Questions to ask before modeling:

- Is response plausibly smooth with dose?
- Does one block behave differently?
- Are there isolated extreme observations?
- Does variance change with dose?
- Are the tested doses dense enough to support curvature?

# 8. Descriptive table

```{r descriptive}
raw_means <- aggregate(yield ~ variety + nitrogen, d, mean)
raw_means
```

These means are not the final inferential result because they ignore the split-plot hierarchy.

# 9. Legacy split-plot design check

Code dose categorically for the classical decomposition.

```{r legacy-data}
d$nitrogen_f <- factor(d$nitrogen)
```

```{r legacy-fit}
legacy <- fit_legacy_splitplot(
  d,
  response = "yield",
  block = "block",
  whole = "variety",
  sub = "nitrogen_f"
)
```

```{r legacy-table}
splitplot_anova(legacy)
```

```{r legacy-ems}
splitplot_ems(d, "block", "variety", "nitrogen_f")
```

The legacy result is used here to verify error strata and design logic.

# 10. Legacy-to-mixed benchmark

```{r legacy-mixed}
if (requireNamespace("lme4", quietly = TRUE)) {
  compare_legacy_mixed(
    d, "yield", "block", "variety", "nitrogen_f"
  )
}
```

# 11. Primary LMM: linear nitrogen trend

Start with the simplest scientifically plausible quantitative response.

```{r linear-fit}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit_linear <- mixed_fit(
    yield ~ variety * nitrogen +
      (1 | block) +
      (1 | block:variety),
    data = d
  )
}
```

# 12. Diagnose the primary model

```{r linear-diagnose}
if (exists("fit_linear")) {
  diag_linear <- mixed_diagnose(fit_linear)
  diag_linear
  mixed_plot(fit_linear, type = "residuals")
}
```

If strong curvature remains, the scientific mean model is incomplete.

# 13. Quadratic nitrogen response

```{r quadratic-fit}
if (requireNamespace("lme4", quietly = TRUE)) {
  fit_quad <- mixed_fit(
    yield ~ variety * poly(nitrogen, 2) +
      (1 | block) +
      (1 | block:variety),
    data = d
  )
}
```

```{r quadratic-diagnose}
if (exists("fit_quad")) {
  diag_quad <- mixed_diagnose(fit_quad)
  diag_quad
  mixed_plot(fit_quad, type = "residuals")
}
```

# 14. Compare mean structures carefully

Because the fixed-effect structures differ, likelihood comparison should use ML if an LRT is desired. For a descriptive comparison, use information criteria while still reviewing residual shape.

```{r ml-fits}
if (requireNamespace("lme4", quietly = TRUE)) {
  linear_ml <- mixed_fit(
    yield ~ variety * nitrogen +
      (1 | block) + (1 | block:variety),
    data = d,
    REML = FALSE
  )

  quad_ml <- mixed_fit(
    yield ~ variety * poly(nitrogen, 2) +
      (1 | block) + (1 | block:variety),
    data = d,
    REML = FALSE
  )
}
```

```{r mean-comparison}
if (exists("linear_ml") && exists("quad_ml")) {
  mixed_compare(
    linear = linear_ml,
    quadratic = quad_ml,
    method = "information"
  )
}
```

# 15. Primary fixed-effect inference

If the quadratic model is retained for scientific and diagnostic reasons, use a small-sample method when appropriate.

```{r kr-anova}
if (
  exists("fit_quad") &&
  requireNamespace("lmerTest", quietly = TRUE) &&
  requireNamespace("pbkrtest", quietly = TRUE)
) {
  mixed_anova(
    fit_quad,
    ddf = "Kenward-Roger"
  )
}
```

# 16. Variety estimated marginal means

```{r variety-emm}
if (
  exists("fit_quad") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  variety_emm <- mixed_emmeans(
    fit_quad,
    specs = ~ variety
  )
  variety_emm
}
```

When a strong interaction is present, averaged variety EMMs require cautious interpretation.

# 17. Planned variety contrasts

```{r variety-contrasts}
if (
  exists("fit_quad") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  mixed_contrasts(
    fit_quad,
    specs = ~ variety,
    method = "pairwise",
    adjust = "tukey"
  )
}
```

# 18. Quantitative nitrogen model helper

Fit the dedicated quadratic quantitative model.

```{r quantitative-helper}
if (requireNamespace("lme4", quietly = TRUE)) {
  qfit <- mixed_quantitative(
    d,
    response = "yield",
    quantitative = "nitrogen",
    qualitative = "variety",
    group = "block",
    degree = 2,
    basis = "polynomial"
  )
}
```

This helper is convenient for trend/curve teaching. For the final split-plot analysis, always verify that the random structure fully represents the randomization required by the study.

# 19. Nitrogen trends

For the linear primary model, marginal slopes are straightforward.

```{r slopes}
if (
  exists("fit_linear") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  slopes <- mixed_trend(
    fit_linear,
    specs = ~ variety,
    var = "nitrogen"
  )
  slopes
}
```

# 20. Response curves

```{r curve-data}
if (
  exists("fit_quad") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  curve_data <- mixed_curve(
    fit_quad,
    variable = "nitrogen",
    values = seq(0, 150, by = 5),
    by = "variety"
  )

  head(curve_data)
}
```

```{r curve-plot, fig.height=5.2}
if (
  exists("fit_quad") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  mixed_plot(
    fit_quad,
    type = "curve",
    x = "nitrogen",
    by = "variety",
    values = seq(0, 150, by = 2.5),
    show_data = TRUE
  )
}
```

**Primary biological interpretation:** describe the dose-response shape, treatment separation over the dose range, and uncertainty. Do not reduce the result to the significance of one polynomial coefficient.

# 21. Model-based optimum

```{r optimum}
if (
  exists("fit_quad") &&
  requireNamespace("emmeans", quietly = TRUE)
) {
  opt <- mixed_optimum(
    fit_quad,
    variable = "nitrogen",
    range = c(0, 150),
    by = "variety",
    objective = "max",
    n = 501
  )

  opt
}
```

The optimum is conditional on the fitted curve and the declared search range. It is not automatically an economic optimum.

# 22. Robust sensitivity analysis

Create the robust counterpart only if the Gaussian LMM is the relevant primary model and `robustlmm` is available.

```{r robust}
if (
  exists("fit_linear") &&
  requireNamespace("robustlmm", quietly = TRUE)
) {
  robust_fit <- mixed_robust(
    yield ~ variety * nitrogen +
      (1 | block) +
      (1 | block:variety),
    data = d
  )

  mixed_compare_robust(
    fit_linear,
    robust_fit
  )
}
```

A material shift should lead to observation/cluster review, not automatic exclusion.

# 23. Contamination screen

```{r contamination}
if (exists("fit_linear")) {
  mixed_contamination(
    fit_linear,
    cluster = "whole_plot",
    threshold = 2.5
  )
}
```

# 24. Bootstrap fixed-effect uncertainty

Use a light vignette-scale bootstrap.

```{r boot-demo}
if (exists("fit_linear")) {
  boot_demo <- mixed_boot_fixed(
    fit_linear,
    B = 49,
    seed = 260818
  )
}
```

Production analysis:

```{r boot-production, eval=FALSE}
boot_full <- mixed_boot_fixed(
  fit_linear,
  B = 4999,
  seed = 260818
)
```

# 25. Bootstrap variance components

```{r boot-var-demo}
if (exists("fit_linear")) {
  boot_var_demo <- mixed_boot_variance(
    fit_linear,
    B = 49,
    seed = 260818
  )
}
```

Variance-component uncertainty can be asymmetric and is often scientifically important in hierarchical experiments.

# 26. Is GAMLSS needed?

Do **not** fit GAMLSS automatically. First ask whether residual scale or shape remains systematically related to treatment or dose after the core mean and covariance structure is adequate.

A sensitivity template:

```{r gamlss-sensitivity}
if (requireNamespace("gamlss", quietly = TRUE)) {
  g_loc <- mixed_gamlss(
    yield ~ variety * nitrogen,
    d,
    family = "NO",
    random = "block"
  )

  g_scale <- mixed_gamlss(
    yield ~ variety * nitrogen,
    d,
    family = "NO",
    sigma.formula = ~ variety,
    random = "block"
  )

  mixed_gamlss_select(
    constant_scale = g_loc,
    variety_scale = g_scale
  )
}
```

This is a distributional sensitivity model, not a replacement for the split-plot randomization analysis unless the full hierarchy is represented appropriately.

# 27. Is a Bayesian model needed?

A Bayesian counterpart is justified when priors, posterior prediction, direct probability statements, or complex hierarchical uncertainty are central to the scientific aim.

```{r bayes-template, eval=FALSE}
priors <- c(
  brms::prior(normal(0, 20), class = "Intercept"),
  brms::prior(normal(0, 5), class = "b"),
  brms::prior(exponential(1), class = "sd"),
  brms::prior(exponential(1), class = "sigma")
)

bayes_fit <- mixed_bayes(
  yield ~ variety * nitrogen +
    (1 | block) +
    (1 | block:variety),
  data = d,
  prior = priors,
  chains = 4,
  iter = 4000,
  seed = 260818
)

mixed_pp_check(bayes_fit)
```

The priors are placeholders for scale-aware refinement, not universal recommendations.

# 28. Fixed-effect table

```{r fixed-table}
if (exists("fit_linear")) {
  mixed_table(
    fit_linear,
    component = "fixed",
    format = "markdown",
    digits = 3,
    caption = "Fixed-effect estimates for the design-correct LMM"
  )
}
```

# 29. Random-effect table

```{r random-table}
if (exists("fit_linear")) {
  mixed_table(
    fit_linear,
    component = "random",
    format = "markdown",
    digits = 3,
    caption = "Random-effect variance components"
  )
}
```

# 30. Diagnostic table

```{r diagnostics-table}
if (exists("diag_linear")) {
  mixed_table(
    diag_linear,
    component = "diagnostics",
    format = "markdown",
    caption = "Diagnostic status"
  )
}
```

# 31. Publication figure set

A compact manuscript set can include:

**Figure A:** observed yield by nitrogen and variety.  
**Figure B:** model-based variety-specific response curves with uncertainty.  
**Figure C:** residual diagnostics.  
**Supplement:** random-effect or bootstrap uncertainty figure.

# 32. Export example

```{r export, eval=FALSE}
mixed_plot(
  fit_quad,
  type = "curve",
  x = "nitrogen",
  by = "variety",
  values = seq(0, 150, by = 2.5),
  show_data = TRUE,
  file = "yield_nitrogen_curve.tiff",
  width = 7,
  height = 5,
  dpi = 600
)
```

# 33. Reproducible report

```{r report}
if (exists("fit_linear")) {
  report_file <- tempfile(fileext = ".md")
  mixed_report(
    fit_linear,
    report_file,
    title = "Split-plot variety × nitrogen analysis"
  )

  readLines(report_file, n = 12)
}
```

# 34. Interpretation framework

A final narrative should follow this sequence.

## Design

State that variety was randomized to whole plots and nitrogen to subplots within blocks.

## Model

State the fixed variety-by-nitrogen mean structure and random block/whole-plot components.

## Diagnostics

Report convergence, singularity, and residual adequacy.

## Inferential result

Describe treatment-specific nitrogen response, uncertainty, and any important interaction.

## Sensitivity

State whether robust or bootstrap analyses changed the substantive conclusion.

## Limits

Do not extrapolate beyond tested nitrogen range or interpret a grid optimum as an economic recommendation without cost/price modeling.

# 35. Decision points for escalation

Escalate only if needed.

| Diagnostic/scientific problem | Extension |
|---|---|
| residual heteroscedasticity | `mixed_heterogeneity()` or distributional scale model |
| serial repeated measurements | `mixed_ar1()` / `mixed_car1()` |
| count response | `mixed_count()` |
| contamination sensitivity | `mixed_robust()` |
| difficult interval uncertainty | `mixed_boot()` |
| changing scale/shape | `mixed_gamlss()` |
| posterior probability/prior question | `mixed_bayes()` |
| nonlinear mechanistic response | `mixed_growth()` / `mixed_dose_response()` |
| field spatial trend | `mixed_spatial_field()` |
| multi-environment adaptation | `mixed_gxe()` |

# 36. What a complete analysis should not look like

Avoid:

1. fitting every available engine;
2. ranking all models by AIC;
3. reporting only the smallest p-value;
4. hiding convergence warnings;
5. deleting influential data without provenance review;
6. presenting EMM letters for quantitative dose as the main result;
7. showing only bars and standard errors;
8. interpreting a Bayesian credible interval as a frequentist CI;
9. treating spatial smoothing as a correction that is always beneficial.

# 37. Reproducibility footer

```{r reproducibility}
utils::sessionInfo()
```

For a manuscript archive, also store:

- source data checksum;
- analysis script;
- seed;
- package versions;
- exported figures/tables;
- model objects when storage permits;
- validation logs.

# 38. Capstone exercises

1. Repeat the workflow with nitrogen categorical and compare the scientific conclusions.
2. Create one missing subplot and explain how the mixed and classical routes differ.
3. Introduce a plausible influential observation and perform robust sensitivity analysis.
4. Bootstrap the variety-by-nitrogen slope and compare with the model-based SE.
5. Write a 250-word results section using the design-model-diagnostic-inference-sensitivity sequence.

# 39. Key take-home messages

- A complete mixed-model analysis begins with randomization, not software.
- Mean structure, covariance, distribution, and robustness are separate modeling questions.
- Quantitative treatments should preserve their scientific scale when possible.
- Advanced modules are escalation paths, not mandatory steps.
- Observed data, uncertainty, diagnostics, and reproducibility should remain visible in the final report.

# 40. Final next step

Use `20-api-example-catalog.Rmd` as the function-level reference after understanding the conceptual workflow developed across the tutorial series.
